Free surface computations in mantle convection models

### 2015-07-24 free-surface-paper.tex
Text

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{amsmath}

@@

Surface topography in simulations of mantle convection and other geodynamic processes is an important observable,
allowing insights into internal density structure, rheology, and gravity perturbations.
Historically, most simulations have been performed with free-slip boundary conditions at the surface, 
with dynamic topography calculated as a postprocessing step.

More recently, the nature of surface boundary conditions have been shown to be important for controlling the 
dynamics of subduction zone modeling.  \citet{schmeling2008benchmark} performed extensive testing on the effect
of a free surface on the sinking of a slab.

@@

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/stability.pdf}
\label{fig:stability}
\caption{Left: stability region for the nonstandard finite difference scheme (green). On the x-axis is the value of the stabilization time, in units of the minimum relaxation time $\tau_{\mathrm{min}}$.  On the y-axis is the value of the timestep, also in units of the minimum relaxation time. For $\tau^*\le \tau_{\mathrm{min}}$ it is unconditionally stable.  Right: stability region for the quasi-implicit scheme (green).  Again, the y axis is in units of the minumum relaxation time.  The x-axis shows the value of the stabilization parameter $\theta$. For $\theta = 1$ it is unconditionally stable. In both cases the stability region for the forward Euler scheme is also shown in blue.}
\end{figure*}

>>>>

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{color, soul}

@@

Surface topography in simulations of mantle convection and other geodynamic processes is an important observable,
allowing insights into Earth's internal density structure, rheology, and geoid perturbations \citep[e.g.][]{richards1984geoid, hager1985lower, baumann2014constraining}.
Historically, most simulations have been performed with free-slip boundary conditions at the surface, 
with dynamic topography calculated as a postprocessing step.

\hl{address pseudo free surface, free surface, sticky air}

More recently, the nature of surface boundary conditions have been shown to be important for controlling the 
dynamics of subduction zone modeling.  In a benchmark study \citet{schmeling2008benchmark} performed extensive testing on the effect
of a free surface on the sinking of a slab. They found that the nature of the free surface had a large effect 
on the dynamics of the slab, affecting both the shape and the timing of sinking. Most of the 
participating codes in that benchmark used the `sticky air' approximation for the free surface, where a low viscosity,
low density layer of fluid (termed `air', though its viscosity is much higher) is placed above the free surface, decoupling 
it from the boundary. Typically a free-slip boundary condition is used above the stiky air.

In particular, \citet{schmeling2008benchmark} found that the specific properties of the sticky air layer 
controlled the shape and timing of the slab, and the viscosity averaging scheme for areas of transitional
composition was extremely important, as the subducting slab entrained significant amounts of the sticky air.
\citet{crameri2012comparison} conducted comparisons between sticky air and true free surface models, 
demonstrating a range of parameters for sticky air which can mitigate some of the difficulties that it 
introduces.

A study by \citet{quinquis2011role} found that free surface boundary conditions have a large effect on 
trench migration in a subduction zone, and \citet{crameri2012free} found that a free surface combined 
with a weak crust is important for producing one-sided subduction zones.

Both true free surface and sticky air calculations have been subject to an instability which has been 
variously termed a ``sloshing,'' or ``druknen sailor'' instability \citep{kaus2010stabilization, duretz2011discretization}. 
This instability, arising from the large density contrast typical at a free surface (compared with the much smaller 
internal density contrasts), severely limits the maximum stable timestep for free surface computations.
Frequently, the maximum stable timestep is several orders of magnitude smaller than that for an otherwise 
equivalent model with free-slip boundary conditions.

Several studies have attempted to alleviate the timestepping requirements imposed by the sloshing 
instability. Since the most expensive part of geodynamic simulations is typically the Stokes solve,
most free surface calculations have preferred to use explicit time stepping methods for the free 
surface. \citet{kaus2010stabilization} proposed a quasi-implicit scheme which modifies the discretized 
Stokes matrix, giving it better stability properties.  \citet{kramer2012implicit} and \citet{furuichi2015implicit}
explored methods for solving the transport of the free surface implicitly.

Here we analyze the problem of the sloshing instability.  We demonstrate the nature of the instability, 
as well as clarifying the effect of quasi-implicit stabilization.

The analysis allows us to propose a new time marching scheme for free surface computations with 
good stability properties. It is an explicit method which requires no modification to the system matrix,
which makes it extremely simple to implement.

Finally, we describe the implementation of a free surface in the mantle convection software \texttt{Aspect}.

@@

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/stability.pdf}
\label{fig:stability}
\caption{Left: stability region for the nonstandard finite difference scheme (green). On the x-axis is the value of the stabilization time, in units of the minimum relaxation time $\tau_{\mathrm{min}}$.  On the y-axis is the value of the timestep, also in units of the minimum relaxation time. For $\tau^*\le 2 \tau_{\mathrm{min}}$ it is unconditionally stable.  Right: stability region for the quasi-implicit scheme (green).  Again, the y axis is in units of the minumum relaxation time.  The x-axis shows the value of the stabilization parameter $\theta$. For $\theta = 1$ it is unconditionally stable. In both cases the stability region for the forward Euler scheme is also shown in blue.}
\end{figure*}
